---
title: "Example 05 - two-way-ANOVA"
format: html
---

```{r setup}
#| echo: false
#| message: false
#| warning: false
 
library(tidyverse)

library(cmdstanr)
library(ggforce)
library(glue)
```

## Simulating data
We are simulating attendance over fifteen weeks for three seminars (A, B, and C ) and for two semesters (W and S). We assume additive effects without an interaction. A-W: 15, B-W: 12, C-W: 13, A-S: 23, B-W: 20, C-W: 21

```{r}
weeks_n <- 30
set.seed(2077)
example05_df <- 
  bind_rows(
    tibble(Seminar = "A", Semester = "W", AttendanceN = rpois(weeks_n, 15)),
    tibble(Seminar = "B", Semester = "W", AttendanceN = rpois(weeks_n, 12)),
    tibble(Seminar = "C", Semester = "W", AttendanceN = rpois(weeks_n, 13)),
    tibble(Seminar = "A", Semester = "S", AttendanceN = rpois(weeks_n, 23)),
    tibble(Seminar = "B", Semester = "S", AttendanceN = rpois(weeks_n, 20)),
    tibble(Seminar = "C", Semester = "S", AttendanceN = rpois(weeks_n, 21))) |>
  mutate(Seminar = factor(Seminar),
         Semester = factor(Semester, levels = c("W", "S")))
         
ggplot(example05_df, aes(x = Seminar, y = AttendanceN)) +
  ggforce::geom_sina() +
  facet_wrap(. ~ Semester)
```

## Model definitions

### Baseline
$$AttendanceN \sim Poisson(\lambda)$$
$$\lambda \sim LogNormal(2.5, 3)$$

```stan
data {
  int<lower=1> N;
  int<lower=1> SeminarsN;
  int<lower=1> SemestersN;
  
  array[N] int<lower=0> AttendanceN;
  array[N] int<lower=1, upper=SeminarsN> Seminar;
  array[N] int<lower=1, upper=SemesterN> Semester;
}

parameters {
  real<lower=0> lambda;
}

model {
  AttendanceN ~ poisson(lambda);
  
  lambda ~ lognormal(2.5, 0.3);
}

generated quantities {
  vector[N] log_lik;
  for(i in 1:N) log_lik[i] = poisson_lpmf(AttendanceN[i] | lambda);
}
```

### Seminar only

$$AttendanceN_i \sim Poisson(\lambda[Seminar_i])$$
$$\lambda[Seminar_i] \sim LogNormal(2.5, 3)$$

```stan
data {
  int<lower=1> N;
  int<lower=1> SeminarsN;
  int<lower=1> SemestersN;
  
  array[N] int<lower=0> AttendanceN;
  array[N] int<lower=1, upper=SeminarsN> Seminar;
  array[N] int<lower=1, upper=SemestersN> Semester;
}

parameters {
  vector<lower=0>[SeminarsN] lambda;
}

model {
  for(i in 1:N) AttendanceN[i] ~ poisson(lambda[Seminar[i]]);
  
  lambda ~ lognormal(2.5, 0.3);
}

generated quantities {
  vector[N] log_lik;
  for(i in 1:N) log_lik[i] = poisson_lpmf(AttendanceN[i] | lambda[Seminar[i]]);
}
```

### Semester only

$$AttendanceN_i \sim Poisson(\lambda[Semester_i])$$
$$\lambda[Semester_i] \sim LogNormal(2.5, 3)$$
```stan
data {
  int<lower=1> N;
  int<lower=1> SeminarsN;
  int<lower=1> SemestersN;
  
  array[N] int<lower=0> AttendanceN;
  array[N] int<lower=1, upper=SeminarsN> Seminar;
  array[N] int<lower=1, upper=SemestersN> Semester;
}

parameters {
  vector<lower=0>[SemestersN] lambda;
}

model {
  for(i in 1:N) AttendanceN[i] ~ poisson(lambda[Semester[i]]);
  
  lambda ~ lognormal(2.5, 0.3);
}

generated quantities {
  vector[N] log_lik;
  for(i in 1:N) log_lik[i] = poisson_lpmf(AttendanceN[i] | lambda[Semester[i]]);
}
```

### Full model

$$AttendanceN_i \sim Poisson(\lambda_i)$$
$$log(\lambda_i) =  \alpha_{AW} + \beta_B\cdot IsB + \beta_C\cdot IsC + \beta_S\cdot IsSummer$$
$$\alpha_{AW} \sim Normal(log(15), 0.5)$$
$$\beta_B, \beta_C, \beta_S \sim Normal(0, 1)$$

```sim
data {
  int<lower=1> N;
  int<lower=1> SeminarsN;
  int<lower=1> SemestersN;
  
  array[N] int<lower=0> AttendanceN;
  array[N] int<lower=1, upper=SeminarsN> Seminar;
  array[N] int<lower=1, upper=SemestersN> Semester;
}

parameters {
  real a_A_W;
  vector[SeminarsN - 1] b_Seminar;
  real b_Semester;
}

transformed parameters {
  matrix[SeminarsN, SemestersN] lambda;
  for(iSeminar in 1:SeminarsN) {
    for(iSemester in 1:SemestersN) {
      if (iSeminar == 1) {
        lambda[iSeminar, iSemester] = exp(a_A_W + b_Semester * (iSemester - 1));
      } else {
        lambda[iSeminar, iSemester] = exp(a_A_W + b_Seminar[iSeminar - 1] + b_Semester * (iSemester - 1));
      }
    }
  }
}

model {
  for(i in 1:N) AttendanceN[i] ~ poisson(lambda[Seminar[i], Semester[i]]);
  
  a_A_W ~ normal(log(15), 0.5);
  b_Seminar ~ normal(0, 1);
  b_Semester ~ normal(0, 1);
}

generated quantities {
  vector[N] log_lik;
  for(i in 1:N) log_lik[i] = poisson_lpmf(AttendanceN[i] | lambda[Seminar[i], Semester[i]]);
}
```
### Interaction model

$$AttendanceN_i \sim Poisson(\lambda[Seminar_i, Semester_i])$$
$$\lambda[Seminar_i, Semester_i] \sim LogNormal(2.5, 3)$$

```stan
data {
  int<lower=1> N;
  int<lower=1> SeminarsN;
  int<lower=1> SemestersN;
  
  array[N] int<lower=0> AttendanceN;
  array[N] int<lower=1, upper=SeminarsN> Seminar;
  array[N] int<lower=1, upper=SemestersN> Semester;
}

parameters {
  matrix<lower=0>[SeminarsN, SemestersN] lambda;
}

model {
  for(i in 1:N) AttendanceN[i] ~ poisson(lambda[Seminar[i], Semester[i]]);
  
  to_vector(lambda) ~ lognormal(2.5, 0.3);
}

generated quantities {
  vector[N] log_lik;
  for(i in 1:N) log_lik[i] = poisson_lpmf(AttendanceN[i] | lambda[Seminar[i], Semester[i]]);
}
```

## Compiling all models
```{r}
example05_modelnames <- c("baseline", "seminar", "semester", "full", "interaction")
example05_models <- map(example05_modelnames, ~cmdstan_model(glue("Stan/example05-{.}.stan")))
names(example05_models) <- example05_modelnames
```

## Preparing the data in Stan format

```{r}
example05_standata <- list(
  N = nrow(example05_df),
  SeminarsN = nlevels(example05_df$Seminar),
  SemestersN = nlevels(example05_df$Semester),
  
  AttendanceN = example05_df$AttendanceN,
  Seminar = as.integer(example05_df$Seminar),
  Semester = as.integer(example05_df$Semester)
)
```

## Sampling all models
```{r}
  example05_fits <- list()
  example05_loos <- list()
  for(amodel in example05_modelnames) {
    example05_fits[[amodel]] <- example05_models[[amodel]]$sample(
      example05_standata,
      chains = 4,
      parallel_chains = parallelly::availableCores())
  example05_loos[[amodel]] <- example05_fits[[amodel]]$loo()
}
```

## Compare models via LOO information criterion

```{r}
loo::loo_compare(example05_loos)
loo::loo_model_weights(example05_loos)
```
