---
title: "Example 04 - two-group pairwise comparison"
format: html
---

```{r setup}
#| echo: false
#| message: false
#| warning: false
 
library(tidyverse)

library(cmdstanr)
library(ggforce)
```

## Simulating data
We are simulating twenty participants assuming 82% correct for full attention condition and 72% for poor attention condition but with (artificially) fixed effect attention.

```{r}
trials_n <- 30
participants_n <- 20
set.seed(2077)

poor_attention_df <-
  tibble(Attention = "Poor",
         Participant = 1:participants_n,
         Ntotal = 30,
         Ncorrect = rbinom(participants_n, size = trials_n, prob = 0.82)) |>
  mutate(Pcorrect = Ncorrect / Ntotal)

full_attention_df <-
  poor_attention_df |>
  mutate(Attention = "Full",
         Scorrect = boot::logit(Pcorrect),
         Pcorrect = boot::inv.logit(Scorrect + 1)) |>
  select(-Scorrect) |>
  mutate(Ncorrect = round(Ntotal * Pcorrect),
         Pcorrect = Ncorrect / Ntotal)

example04_df <- 
  bind_rows(poor_attention_df, full_attention_df) |>
  mutate(Attention = factor(Attention, levels = c("Poor", "Full")))
```
  

## Model definition: difference as parameter, central parametrization

$$Ncorrect_i \sim Binomial(Ntotal_i, p_i)$$
$$logit(p_i) = \alpha_{poor}[Participant_i] + \beta_{A} IsFull_i$$
$$\mu_{\alpha_{poor}} \sim Normal(logit(0.75), 0.5)$$
$$\alpha_{poor}[Participant_i] \sim Normal(\mu_{\alpha_{poor}}, \sigma_{Participant})$$
$$\sigma_{Participant} \sim Exponential(1)$$
$$\beta_A \sim Normal(0, 1)$$

```stan
data {
  int<lower=1> N;
  int<lower=1> AttentionN;
  int<lower=1> ParticipantsN;
  
  array[N] int<lower=1> Ntotal;
  array[N] int<lower=0> Ncorrect;
  array[N] int<lower=1, upper=AttentionN> Attention;
  array[N] int<lower=1, upper=ParticipantsN> Participant;
}

transformed data {
  array[N] int IsFull;
  for(i in 1:N) IsFull[i] = Attention[i] - 1;
}

parameters {
  real mu_a_poor;                    // group-level average for poor attention condition
  real<lower=0> sigma_Participant;   // variability of population
  vector[ParticipantsN] a_poor;      // poor attention condition for individual participants
  real b_A;                          // group-level difference between attention conditions
}

transformed parameters {
  vector[N] p;
  for(i in 1:N) p[i] = inv_logit(a_poor[Participant[i]] + b_A * IsFull[i]);
}

model {
  Ncorrect ~ binomial(Ntotal, p);

  mu_a_poor ~ normal(logit(0.75), 1);
  sigma_Participant ~ exponential(1);
  a_poor ~ normal(mu_a_poor, sigma_Participant);
  b_A ~ normal(0, 1);
}

generated quantities {
  // probability of correct response at group level
  vector[2] pCorrect;
  pCorrect[1] = inv_logit(mu_a_poor);
  pCorrect[2] = inv_logit(mu_a_poor + b_A);

  // difference between conditions in probability space  
  real dA = pCorrect[2] - pCorrect[1];
}
```

Compiling the model
```{r}
example04_model <- cmdstan_model("Stan/example04-via-difference-centered.stan")
```

Preparing the data in Stan format

```{r}
example04_standata <- list(
  N = nrow(example04_df),
  AttentionN = nlevels(example04_df$Attention),
  ParticipantsN = max(example04_df$Participant),
  
  Ntotal = example04_df$Ntotal,
  Ncorrect = example04_df$Ncorrect,
  Attention = as.integer(example04_df$Attention),
  Participant = example04_df$Participant
)
```

Sampling the model
```{r}
# sample the model
example04_fit <- example04_model$sample(
  example04_standata,
  chains = 4,
  parallel_chains = parallelly::availableCores())
```

## Same model, non-centered parameterization

$$Ncorrect_i \sim Binomial(Ntotal_i, p_i)$$
$$logit(p_i) = \alpha_{poor}[Participant_i] + \beta_{A} IsFull_i$$
$$\mu_{\alpha_{poor}} \sim Normal(logit(0.75), 0.5)$$

$$z_{poor} \sim Normal(0, 1)$$
$$\sigma_{Participant} \sim Exponential(1)$$
$$\alpha_{poor} =  \mu_{\alpha_{poor}} + \sigma_{Participant}) \cdot z_{poor}$$
$$\beta_A \sim Normal(0, 1)$$
```stan
data {
  int<lower=1> N;
  int<lower=1> AttentionN;
  int<lower=1> ParticipantsN;
  
  array[N] int<lower=1> Ntotal;
  array[N] int<lower=0> Ncorrect;
  array[N] int<lower=1, upper=AttentionN> Attention;
  array[N] int<lower=1, upper=ParticipantsN> Participant;
}

transformed data {
  array[N] int IsFull;
  // alternative via for loop
  for(i in 1:N) IsFull[i] = Attention[i] - 1; // Or Attention[i] == 2
}

parameters {
  real mu_a_poor;                    // group-level average for poor attention condition
  real<lower=0> sigma_Participant;   // variability of population
  vector[ParticipantsN] z_poor;      // poor attention condition for individual participants
  real b_A;                          // group-level difference between attention conditions
}

transformed parameters {
  // poor attention condition for individual participants
  vector[ParticipantsN] a_poor = mu_a_poor + sigma_Participant * z_poor;
      
  vector[N] p;
  for(i in 1:N) p[i] = inv_logit(a_poor[Participant[i]] + b_A * IsFull[i]);
}

model {
  Ncorrect ~ binomial(Ntotal, p);

  mu_a_poor ~ normal(logit(0.75), 1);
  sigma_Participant ~ exponential(1);
  z_poor ~ normal(0, 1);
  b_A ~ normal(0, 1);
}

generated quantities {
  // probability of correct response at group level
  vector[2] pCorrect;
  pCorrect[1] = inv_logit(mu_a_poor);
  pCorrect[2] = inv_logit(mu_a_poor + b_A);

  // difference between conditions in probability space  
  real dA = pCorrect[2] - pCorrect[1];
}
```

Compiling the model
```{r}
example04_model <- cmdstan_model("Stan/example04-via-difference-noncentered.stan")
```

Sampling the model
```{r}
# sample the model
example04_fit <- example04_model$sample(
  example04_standata,
  chains = 4,
  parallel_chains = parallelly::availableCores())
```

Extracting samples
```{r}
example04_samples <- example04_fit$draws(format = "df") |> as_tibble()
```

Plot posterior distribution for estimated proportion of correct responses per condition

```{r}
# getting probability of correct response as a long table
posterior_pcorrect <-
  example04_samples |>
  select(.draw, starts_with('pCorrect')) |>
  pivot_longer(starts_with('pCorrect'), names_to = "Attention", values_to = "Pcorrect") |>
  mutate(Attention = factor(Attention, labels = levels(example04_df$Attention)))

ggplot(posterior_pcorrect, aes(x = Attention, y = Pcorrect)) +
  geom_violin() +
  ggforce::geom_sina(data = example04_df)
```

Plot posterior distribution for estimated difference in log-odds (b_A)

```{r}
# compute 97% credible interval + median
b_A_quantiles <- quantile(example04_samples$b_A, c((1-0.97)/2, 0.5, 1-(1-0.97)/2))

ggplot(example04_samples, aes(x = b_A)) +
  geom_histogram(bins = 100) +
  geom_vline(xintercept = b_A_quantiles, linetype = "dashed") +
  geom_vline(xintercept = 0, color = "red") +
  scale_x_continuous(name = "Pcorrect") +
  labs(subtitle = sprintf("P(b_A>0)=%.1f%%", 100 * mean(example04_samples$b_A > 0)))
```

Plot posterior distribution for estimated difference in units of probability

```{r}
# compute 97% credible interval + median
dA_quantiles <- quantile(example04_samples$dA, c((1-0.97)/2, 0.5, 1-(1-0.97)/2))

ggplot(example04_samples, aes(x = dA)) +
  geom_histogram(bins = 100) +
  geom_vline(xintercept = dA_quantiles, linetype = "dashed") +
  geom_vline(xintercept = 0, color = "red") +
  scale_x_continuous(name = "Pcorrect") +
  labs(subtitle = sprintf("P(F>P)=%.1f%%", 100 * mean(example04_samples$dA > 0)))
```


## Same model but via indexing approach
$$Ncorrect_i \sim Binomial(Ntotal_i, p_i)$$
$$logit(p_i) = \alpha[Attention_i, Participant_i]$$
$$\mu_{\alpha} \sim Normal(logit(0.75), 0.5)$$
$$z_{poor} \sim Normal(0, 1)$$
$$\sigma_{Participant} \sim Exponential(1)$$
$$\alpha[Attention_i, Participant_i] =  \mu_{\alpha}[Attention_i] + \sigma_{Participant}) \cdot z[Participant_i]$$


```stan
data {
  int<lower=1> N;
  int<lower=1> AttentionN;
  int<lower=1> ParticipantsN;
  
  array[N] int<lower=1> Ntotal;
  array[N] int<lower=0> Ncorrect;
  array[N] int<lower=1, upper=AttentionN> Attention;
  array[N] int<lower=1, upper=ParticipantsN> Participant;
}

parameters {
  vector[AttentionN] mu_a;           // group-level average for poor attention condition
  real<lower=0> sigma_Participant;   // variability of population
  vector[ParticipantsN] z;      // poor attention condition for individual participants
  real b_A;                          // group-level difference between attention conditions
}

transformed parameters {
  // poor attention condition for individual participants
  matrix[AttentionN, ParticipantsN] a;
  for(iA in 1:AttentionN){
    for(iP in 1:ParticipantsN){
      a[iA, iP] = mu_a[iA] + sigma_Participant * z[iP];
    }
  }
      
  vector[N] p;
  for(i in 1:N) p[i] = inv_logit(a[Attention[i], Participant[i]]);
}

model {
  Ncorrect ~ binomial(Ntotal, p);

  mu_a ~ normal(logit(0.75), 1);
  sigma_Participant ~ exponential(1);
  z ~ normal(0, 1);
}

generated quantities {
  // probability of correct response at group level
  vector[2] pCorrect = inv_logit(mu_a);

  // difference between conditions in probability space  
  real dA = pCorrect[2] - pCorrect[1];
}
```

Compiling the model
```{r}
example04_indexing_model <- cmdstan_model("Stan/example04-via-indexing-noncentered.stan")
```
Sampling
```{r}
# sample the model
example04_indexing_fit <- example04_indexing_model$sample(
  example04_standata,
  chains = 4,
  parallel_chains = parallelly::availableCores())
```

Extracting samples
```{r}
example04_indexing_samples <- example04_indexing_fit$draws(format = "df") |> as_tibble()
```

Plot posterior distribution for estimated proportion of correct responses per condition

```{r}
# getting probability of correct response as a long table
posterior_pcorrect <-
  example04_indexing_samples |>
  select(.draw, starts_with('pCorrect')) |>
  pivot_longer(starts_with('pCorrect'), names_to = "Attention", values_to = "Pcorrect") |>
  mutate(Attention = factor(Attention, labels = levels(example04_df$Attention)))

ggplot(posterior_pcorrect, aes(x = Attention, y = Pcorrect)) +
  geom_violin() +
  ggforce::geom_sina(data = example04_df)
```

Plot posterior distribution for estimated difference in units of probability

```{r}
# compute 97% credible interval + median
dA_quantiles <- quantile(example04_indexing_samples$dA, c((1-0.97)/2, 0.5, 1-(1-0.97)/2))

ggplot(example04_indexing_samples, aes(x = dA)) +
  geom_histogram(bins = 100) +
  geom_vline(xintercept = dA_quantiles, linetype = "dashed") +
  geom_vline(xintercept = 0, color = "red") +
  scale_x_continuous(name = "Pcorrect") +
  labs(subtitle = sprintf("P(F>P)=%.1f%%", 100 * mean(example04_indexing_samples$dA > 0)))
```
