---
title: "Example 03 - two-group comparison"
format: html
---

```{r setup}
#| echo: false
#| message: false
#| warning: false
 
library(tidyverse)

library(cmdstanr)
```

## Simulating data
We are simulating thirty trials assuming 82% correct for full attention condition and 72% for poor attention condition.

```{r}
trials_n <- 30
set.seed(2077)
example03_df <- 
  tibble(
    Attention = factor(rep(c("Full", "Poor"), each=trials_n), levels = c("Poor", "Full")),
    Correct = c(rbinom(trials_n, 1, 0.82), rbinom(trials_n, 1, 0.72)))

example03_avg <-
  example03_df |>
  group_by(Attention) |>
  summarise(Pcorrect = mean(Correct), .groups = "drop")

example03_avg
```

## Model definition: difference as parameter

$$Correct \sim Bernoulli(p_i)$$
$$logit(p_i) = \alpha_{poor} + \beta_{A} IsFull_i$$

$$\alpha_{poor} \sim Normal(logit(0.75), 0.5)$$
$$\beta_A \sim Normal(0, 1)$$

```stan
data {
  int<lower=1> N;
  int<lower=1> AttentionN;
  
  array[N] int<lower=0, upper=1> Correct;
  array[N] int<lower=1, upper=AttentionN> Attention;
}

transformed data {
  array[N] int IsFull = to_vector(Attention) - 1;
  // alternative via for loop
  // for(i in 1:N) IsStats[i] = Seminar[i] - 1;
}

parameters {
  real a_poor;
  real b_A;
}

transformed parameters {
  vector[N] p;
  for(i in 1:N) {
    // logit link, i.e., logit(p[i]) = a_poor + b_A * IsFull[i]
    p[i] = inv_logit(a_poor + b_A * IsFull[i]);
  }
}

model {
  Correct[i] ~ bernoulli(p);

  a_poor ~ normal(logit(0.75), 0.5);
  b_A ~ normal(0, 1);
}

generated quantities {
  // probability of correct response
  vector[2] pCorrect;
  pCorrect[1] = inv_logit(a_poor);
  pCorrect[2] = inv_logit(a_poor + b_A);

  // difference between conditions in probability space  
  real dA = pCorrect[2] - pCorrect[1];
}
```

## Compiling the model
```{r}
example03_model <- cmdstan_model("Stan/example03-via-difference.stan")
```

## Preparing the data in Stan format

```{r}
example03_standata <- list(
  N = nrow(example03_df),
  AttentionN = nlevels(example03_df$Attention),
  
  Correct = example03_df$Correct,
  Attention = as.integer(example03_df$Attention)
)
```

## Sampling the model
```{r}
# sample the model
example03_fit <- example03_model$sample(
  example03_standata,
  chains = 4,
  parallel_chains = parallelly::availableCores())
```

## Extracting samples
```{r}
example03_samples <- example03_fit$draws(format = "df") |> as_tibble()
```

## Plot posterior distribution for estimated proportion of correct responses per condition

```{r}
# getting probability of correct response as a long table
posterior_pcorrect <-
  example03_samples |>
  select(.draw, starts_with('pCorrect')) |>
  pivot_longer(starts_with('pCorrect'), names_to = "Attention", values_to = "Pcorrect") |>
  mutate(Attention = factor(Attention, labels = levels(example03_df$Attention)))

ggplot(posterior_pcorrect, aes(x = Attention, y = Pcorrect)) +
  geom_violin() +
  geom_errorbar(data = example03_avg, aes(ymin = Pcorrect, ymax = Pcorrect))
```

## Plot posterior distribution for estimated difference in log-odds (b_A)

```{r}
# compute 97% credible interval + median
b_A_quantiles <- quantile(example03_samples$b_A, c((1-0.97)/2, 0.5, 1-(1-0.97)/2))

ggplot(example03_samples, aes(x = b_A)) +
  geom_histogram(bins = 100) +
  geom_vline(xintercept = b_A_quantiles, linetype = "dashed") +
  geom_vline(xintercept = 0, color = "red") +
  scale_x_continuous(name = "Pcorrect") +
  labs(subtitle = sprintf("P(b_A>0)=%.1f%%", 100 * mean(example03_samples$b_A > 0)))
```

## Plot posterior distribution for estimated difference in units of probability

```{r}
# compute 97% credible interval + median
dA_quantiles <- quantile(example03_samples$dA, c((1-0.97)/2, 0.5, 1-(1-0.97)/2))

ggplot(example03_samples, aes(x = dA)) +
  geom_histogram(bins = 100) +
  geom_vline(xintercept = dA_quantiles, linetype = "dashed") +
  geom_vline(xintercept = 0, color = "red") +
  scale_x_continuous(name = "Pcorrect") +
  labs(subtitle = sprintf("P(F>P)=%.1f%%", 100 * mean(example03_samples$dA > 0)))
```


## Same model but via indexing approach

$$Correct \sim Bernoulli(p_i)$$
$$p_i = p_{correct}[Attention_i]$$

$$p_{correct} \sim BetaProportion(0.75, 5)$$
```stan
data {
  int<lower=1> N;
  int<lower=1> AttentionN;
  
  array[N] int<lower=0, upper=1> Correct;
  array[N] int<lower=1, upper=AttentionN> Attention;
}

parameters {
  vector[AttentionN] pCorrect;
}

transformed parameters {
  vector[N] p;
  for(i in 1:N) p[i] = pCorrect[Attention[i]];
}

model {
  Correct ~ bernoulli(p);

  pCorrect ~ beta_proportion(0.75, 5);
}

generated quantities {
  // difference between conditions in probability space  
  real dA = pCorrect[2] - pCorrect[1];
}
```

Compiling the model
```{r}
example03_indexing_model <- cmdstan_model("Stan/example03-via-indexing.stan")
```

Sampling
```{r}
# sample the model
example03_indexing_fit <- example03_indexing_model$sample(
  example03_standata,
  chains = 4,
  parallel_chains = parallelly::availableCores())
```

Extracting samples
```{r}
example03_indexing_samples <- example03_indexing_fit$draws(format = "df") |> as_tibble()
```

Plot posterior distribution for estimated proportion of correct responses per condition

```{r}
# getting probability of correct response as a long table
posterior_pcorrect <-
  example03_indexing_samples |>
  select(.draw, starts_with('pCorrect')) |>
  pivot_longer(starts_with('pCorrect'), names_to = "Attention", values_to = "Pcorrect") |>
  mutate(Attention = factor(Attention, labels = levels(example03_df$Attention)))

ggplot(posterior_pcorrect, aes(x = Attention, y = Pcorrect)) +
  geom_violin() +
  geom_errorbar(data = example03_avg, aes(ymin = Pcorrect, ymax = Pcorrect))
```

Plot posterior distribution for estimated difference in units of probability

```{r}
# compute 97% credible interval + median
dA_quantiles <- quantile(example03_indexing_samples$dA, c((1-0.97)/2, 0.5, 1-(1-0.97)/2))

ggplot(example03_indexing_samples, aes(x = dA)) +
  geom_histogram(bins = 100) +
  geom_vline(xintercept = dA_quantiles, linetype = "dashed") +
  geom_vline(xintercept = 0, color = "red") +
  scale_x_continuous(name = "Pcorrect") +
  labs(subtitle = sprintf("P(F>P)=%.1f%%", 100 * mean(example03_indexing_samples$dA > 0)))
```
