---
title: "Exercise 01 - Poisson counts"
format: html
---

```{r setup}
#| echo: false
#| message: false
#| warning: false
 
library(tidyverse)

library(cmdstanr)
```

## Simulating data
We are simulating 15 presence counts assuming 12.5 average attendance.
```{r}
set.seed(1977)
exercise01_df <- tibble(PresentN = rpois(15, 12.5))
```

## Model

Create a model in Stan for count data. The distribution you sample data from is [poisson(lambda)](https://mc-stan.org/docs/functions-reference/unbounded_discrete_distributions.html#poisson). The parameter $\lambda$ can be sampled from prior distribution $lognormal(2.5, 0.35)$.


```stan

```

## Compiling the model
```{r}
exercise01_model <- cmdstan_model("Stan/exercise01.stan")
```

## Preparing the data in Stan format

```{r}
exercise01_standata <- list(
  N = nrow(exercise01_df),
  PresentN = exercise01_df$PresentN
)
```

## Sampling the model
```{r}
# sample the model
exercise01_fit <- exercise01_model$sample(
  exercise01_standata,
  chains = 4,
  parallel_chains = parallelly::availableCores())
```

## Extracting samples
```{r}
exercise01_samples <- exercise01_fit$draws(format = "df") |> as_tibble()
```

## Plot posterior distribution for attendance

```{r}
# compute 97% credible interval + median
lambda_quantiles <- quantile(exercise01_samples$lambda, c((1-0.97)/2, 0.5, 1-(1-0.97)/2))

ggplot(exercise01_samples, aes(x = lambda)) +
  geom_histogram(bins = 100) +
  geom_vline(xintercept = lambda_quantiles, linetype = "dashed") +
  geom_vline(xintercept = mean(exercise01_df$PresentN), color = "blue") +
  scale_x_continuous(name = "Presence") 
```

## Compute probablity that performance is above 12

```{r}
mean(exercise01_samples$lambda > 12)
```
