---
title: "Exercise 02 - Normal height"
format: html
---

```{r setup}
#| echo: false
#| message: false
#| warning: false
 
library(tidyverse)

library(cmdstanr)
```

## Simulating data
We are simulating 23 average heights.
```{r}
set.seed(1977)
exercise02_df <- tibble(Height = rnorm(23, 170, 10))
```

## Model

Create a model in Stan for height data. We can assume that data can be described by a normal distribution: [normal(mu, sigma)](https://mc-stan.org/docs/functions-reference/unbounded_continuous_distributions.html#normal-distribution).

$$Height \sim Normal(\mu, \sigma)$$
$$\mu \sim Normal(178, 5)$$
$$\sigma \sim Exponential(1)$$

```stan
data {
  int<lower=1> N;
  array[N] real Height;
}
parameters {
  real mu;
  real<lower=0> sigma;
}
model {
  Height ~ normal(mu, sigma);   
  
  mu ~ normal(178, 5);
  sigma ~ exponential(1);
}
```

## Compiling the model
```{r}
exercise02_model <- cmdstan_model("Stan/exercise02.stan")
```

## Preparing the data in Stan format

```{r}
exercise02_standata <- list(
  N = nrow(exercise02_df),
  Height = exercise02_df$Height
)
```

## Sampling the model
```{r}
# sample the model
exercise02_fit <- exercise02_model$sample(
  exercise02_standata,
  chains = 4,
  parallel_chains = parallelly::availableCores())
```

## Extracting samples
```{r}
exercise02_samples <- exercise02_fit$draws(format = "df") |> as_tibble()
```

## Plot posterior distribution for attendance

```{r}
# compute 97% credible interval + median
mu_quantiles <- quantile(exercise02_samples$mu, c((1-0.97)/2, 0.5, 1-(1-0.97)/2))

ggplot(exercise02_samples, aes(x = mu)) +
  geom_histogram(bins = 100) +
  geom_vline(xintercept = mu_quantiles, linetype = "dashed") +
  geom_vline(xintercept = mean(exercise02_df$Height), color = "blue") +
  scale_x_continuous(name = "Presence") 
```

## Compute probability that average group height is above average height of 178

```{r}
mean(exercise02_samples$mu > 178)
```
